Integration of national demographic-disturbance relationships into local caribou population viability projections can reduce uncertainty and inform monitoring decisions

### 2024-08-26 analysis/paper/paper.Rmd
minor text edits

To derive a composition survey bias correction term we assume each group of animals in a calf:cow composition survey contains one or more collared adult females ($T$), and may also include: uncollared adult females misidentified as young bulls or unknown sex ($U$); correctly identified uncollared adult females ($V$); young bulls correctly identified as male or unknown sex ($O$); young bulls misidentified as uncollared adult females ($P$); observed calves ($J$); and unobserved calves ($K$). The apparent number of adult females in the group is $T+V+P=Tw$, where $w$ is the observed ratio of the apparent number of uncollared adult females to collared animals in the recruitment survey. The ratio of young bulls to uncollared adult females in the group is
$q = (P+O)/(U+V).$ Assuming an equal probability $u$ of misidentifying young bulls as adult females and vice versa, we get $V=(U+V)(1-u)$ and $P=(O+P)u$. Given a probability $z$ of missing calves, we get $J=(J+K)(1-z)$. Given these assumptions and definitions, the sex and bias-corrected recruitment rate $X$ can be written as function of the observed calf:cow ratio $R$, the cow multiplier $w$, the ratio of young bulls to adult females $q$, and the misidentification probabilities $u$ and $z$ (see Supplement B for details). For simplicity, we combine the probabilities into a bias correction term $c$ that is integrated into the Bayesian population model and simulations (below): 
$$c=\frac{w(1+qu-u)}{(w+qu-u)(1-z)}; X=cR/2$$

@@

$\phi^R \sim \text{Normal}(19.862,2.229)$ and
$\phi^S \sim \text{Normal}(63.733,8.311)$ are precisions of the Beta
distributed errors [@ferrari_beta_2004]. At the beginning of a
simulation for an example population, regression coefficient values are
sampled from Gaussian distributions (see Table \@ref(tab:priors)) and
the population is assigned to quantiles of the Beta error distributions
for survival and recruitment. The population remains in these quantiles
as disturbance changes over time, so there is substantial persistent
variation in recruitment and survival among example among populations
(Figure \@ref(fig:exampleTrajectories)). To ensure results are comparable across monitoring scenarios and that populations remain extant through the monitoring period, all simulations are started with $N_0=5000$ animals 24 years prior to the start of the projection period.

@@

Following others [@eacker_web-based_2019;@lamb_effectiveness_2024], Kaplan-Meier
estimates of observed adult female survival probability $\hat{S}_t$ and
associated precisions $\tau_t$ are estimated from known-fate radio
collar data using the _survival_ R package [@therneau_modeling_2000], with
precisions for years with no observed mortality estimated using a
separate Bayesian model. Observed estimated survival is a modelled as a Gaussian
distributed function of the true survival probability, truncated between 0 and 1, to allow sharing survival information across years with a random effect [@eacker_web-based_2019,@lamb_effectiveness_2024]:
\begin{equation}
\hat{S_t} \sim \text{Normal}(S_t,1/\tau_t)T(0,1).
  (\#eq:hatSEst)
\end{equation}

@@

knitr::kable(tbl,caption="Prior means and standard deviations of Gaussian distributed survival and reproduction model parameters. Standard deviations are calibrated so that the 95\\% prior predictive intervals for survival and recruitment from the Bayesian model match the
the 95% predictive intervals from Johnson et al's regression model (Supplement C).")
```

@@

As suggested by Dunham et al. [-@dunham_strategic_2023] we also calculate the expected value of perfect information (EVPI) and the expected value of sample information (EVSI) [@runge_which_2011;@yokota_value_2004]. We assign a value or utility of $\nu_{\alpha,\psi}=1$ when the true state of the population $\psi$ matches the assigned a status $\alpha$, and value of $\nu_{\alpha,\psi}=0$ when the status assessment is incorrect. This is "the value ascribed by the decision-maker to the outcome, and is thus the measure of management performance" [@runge_which_2011]. The expected value of current information is the average utility of decisions informed by knowledge of the disturbance scenario $\tau$ (Fig \@ref(fig:figDisturbance)) and priors from the national model:
\begin{equation}
EV_{current} = max_{\alpha}\sum_\psi[P(\psi|\tau)\nu_{\alpha,\psi}],
  (\#eq:EVcurrent)
\end{equation}
where $P(\psi|\tau)$ is the prior probability that the true state of the population is $\psi$, and each population is assigned a status $\alpha$ informed by the prior. The expected value under certainty (i.e. the expected value of a decision when the true state of the population is known) is $EV_{certainty}=1$. The difference between $EV_{certainty}$ and $EV_{current}$ gives EVPI, a measure of the maximum possible benefit that could theoretically be gained if the system were perfectly known.

EVSI measures expected improvement from a particular sampling strategy as the difference between the expected value with and without additional sample information: 
\begin{equation}
EVSI_{\tau} = E_x[max_{\alpha}\sum_\psi[P(\psi|x,\tau)\nu_{\alpha,\psi}]] - EV_{current},
  (\#eq:EVSI)
\end{equation}
where $x$ is the sample information from a particular example trajectory. In this study, we treat simulated observations from 500 example trajectories for each disturbance and monitoring scenario $\tau$ (Figure \@ref(fig:figDisturbance)) as sample from the distribution of possible values of $x$, and assume each of these sample results is equally probable. $P(\psi|x,\tau)$ is the posterior probability that the true state of the population is $\psi$, and each population is assigned a status $\alpha$ informed by the sample information.

@@

Our analysis becomes more informative when we examine the probability of correctly assessing population status (Fig \@ref(fig:figStatusErrors)) and the expected value of information (Figs \@ref(fig:figEVSI) and E.1). When anthropogenic disturbance is high (right panels of Figs \@ref(fig:figStatusErrors), \@ref(fig:figEVSI), and E.1)), projections of population status informed only by priors from the national demographic-disturbance relationships are accurate, and adding additional local monitoring data does not improve accuracy. Interestingly, when the projection period is long and anthropogenic disturbance is high, we see a negative EVSI (lower right panel of Fig \@ref(fig:figEVSI)), and we do not see an associated decrease in the probability of correctly assessing population status (lower right panel of Fig \@ref(fig:figStatusErrors)). This means that the long-term projections informed by local monitoring data are on average less precise but not less accurate, which can happen when there is very small prior probability that population is viable (orange band in the top right panel of Fig \@ref(fig:examplePredictions) falls below 1). At lower levels of disturbance, local monitoring can reduce uncertainty, but there are diminishing returns as the number of collars deployed or the number of years of monitoring increases (Figs \@ref(fig:figStatusErrors) and \@ref(fig:figEVSI)). The shapes of the relationships between monitoring effort and the probability of correct status assessment depend on how anthropogenic disturbance is changing over time, but there are diminishing benefits of increased monitoring regardless of whether disturbance is increasing or decreasing (supplement D). Adding unknown bias increases uncertainty, but these increases are not large for the set of parameters we investigate (supplement F).

@@

```{r figStatusErrors, fig.cap="The probability of a correct status assessment varies with the amount of anthropogenic disturbance (columns), the number of years projected (rows), and monitoring effort (see Figure \\@ref(fig:figDisturbance) for scenarios). When disturbance is high the probability of correct status assessment is high even without local monitoring. Where disturbance is low or moderate, increasing the number of years of monitoring and the number of collars reduces uncertainty, but there are diminishing returns on additional monitoring (number of years or number of collars) as the number of years or collars increases. In these examples the composition cow multiplier $w$ is 6; varying $w$ does not alter qualitative conclusions (supplements D and E).", cache=TRUE}
knitr::include_graphics(here::here("figs/forPaper/powercmult6ay3aSf1.png"))
```

@@

There are a variety of possible actions intended to improve survival and reproduction of populations that are not otherwise self-sustaining [@mcnay_demographic_2022; @serrouya_predicting_2020]. The simple models in this paper do not include these actions, and cannot be used to project outcomes for managed populations. Our intention is to reduce uncertainty that can be used to delay conservation action; there is nothing in our arguments, models or results that can or should be used to discourage much needed monitoring of the effects of conservation actions. Even when no additional additional monitoring is required for assessing self-sustainability, there are a variety of reasons to continue monitoring highly disturbed boreal caribou populations. Our point is that action is also required, and lack of monitoring should not be used as a reason to delay conservation measures for highly disturbed populations.

@@

Another novel aspect of this work is a term that allows for bias associated with misclassification of adults and missing calves in composition surveys [@ellington_divergent_2020; @decesare_estimating_2012]. Lacking information about the true magnitude of this bias, we examined sensitivity to a range of assumed values, and found that adding unknown bias did increase uncertainty, but these increases were not large (supplement E). A key reason for this result is that we assumed the same range of values for the probability of misclassifying adults and the probability of missing calves. Presuming that surveyed groups contain more cows than bulls, and that there is an equal probability of misclassifying cows and young bulls and vice versa, then misclassification errors cause high estimated recruitment. Missing calves causes low estimated recruitment, and together these errors tend to cancel one another out. We did not investigate sensitivity to the assumption that new recruits are 50% females, but we note that this is another possible source of error that would most likely lead to low recruitment estimates.

@@

The main objective of Johnson et al's [-@johnson_science_2020;@eccc_scientific_2011] national analysis was to inform a national disturbance policy that applies to all ranges except SK1, and to inform a different policy for SK1 [@eccc_scientific_2011;@eccc_boreal_2019]. A hierarchical analysis [@leasure_hierarchical_2019] would be better suited for estimating variation in demographic model parameters among populations, and distinguishing that from variation among years. However, data owners have not authorized the use of the data for that purpose. In order to proceed with the available published information we constrained our example trajectory simulations by assuming that populations remain in their quantile, so populations that are lower than higher than average remain lower or higher than average over time. The Bayesian model is not similarly constrained, and can accommodate evidence of a shift from unexpectedly low to unexpectedly low demographic rates over time, or vice versa. Switching from low to high each year is unlikely, but there is evidence of variation in the slope demographic-disturbance relationships among populations [@rudolph_demographic_2017] so it is important that the Bayesian model allows for a greater range of possibilities than our more constrained simulated example trajectories. Although our simulated trajectories do not include all plausible possibilities we believe they are sufficient to demonstrate how a Bayesian model can be used to integrate local data with national priors, and to show how monitoring requirements depend on landscape condition.

>>>>

To derive a composition survey bias correction term we assume each group of animals in a calf:cow composition survey contains one or more collared adult females ($T$), and may also include: uncollared adult females misidentified as young bulls or unknown sex ($U$); correctly identified uncollared adult females ($V$); young bulls correctly identified as male or unknown sex ($O$); young bulls misidentified as uncollared adult females ($P$); observed calves ($J$); and unobserved calves ($K$). The apparent number of adult females in the group is $T+V+P=Tw$, where $w$ is the observed ratio of the apparent number of uncollared adult females to collared animals in the recruitment survey. The ratio of young bulls to uncollared adult females in the group is
$q = (P+O)/(U+V).$ Assuming an equal probability $u$ of misidentifying young bulls as adult females and vice versa, we get $V=(U+V)(1-u)$ and $P=(O+P)u$. Given a probability $z$ of missing calves, we get $J=(J+K)(1-z)$. Given these assumptions and definitions, the sex and bias-corrected recruitment rate $X$ can be written as a function of the observed calf:cow ratio $R$, the cow multiplier $w$, the ratio of young bulls to adult females $q$, and the misidentification probabilities $u$ and $z$ (see Supplement B for details). For simplicity, we combine the probabilities into a bias correction term $c$ that is integrated into the Bayesian population model and simulations (below): 
$$c=\frac{w(1+qu-u)}{(w+qu-u)(1-z)}; X=cR/2$$

@@

$\phi^R \sim \text{Normal}(19.862,2.229)$ and
$\phi^S \sim \text{Normal}(63.733,8.311)$ are precisions of the Beta
distributed errors [@ferrari_beta_2004]. At the beginning of a
simulation for an example population, regression coefficient values are
sampled from Gaussian distributions (see Table \@ref(tab:priors)) and
the population is assigned to quantiles of the Beta error distributions
for survival and recruitment. The population remains in these quantiles
as disturbance changes over time, so there is substantial persistent
variation in recruitment and survival among example populations
(Figure \@ref(fig:exampleTrajectories)). To ensure results are comparable across monitoring scenarios and that populations remain extant through the monitoring period, all simulations are started with $N_0=5000$ animals 24 years prior to the start of the projection period.

@@

Following others [@eacker_web-based_2019;@lamb_effectiveness_2024], Kaplan-Meier
estimates of observed adult female survival probability $\hat{S}_t$ and
associated precisions $\tau_t$ are estimated from known-fate radio
collar data using the _survival_ R package [@therneau_modeling_2000], with
precisions for years with no observed mortality estimated using a
separate Bayesian model. Observed estimated survival is modelled as a Gaussian
distributed function of the true survival probability, truncated between 0 and 1, to allow sharing survival information across years with a random effect [@eacker_web-based_2019,@lamb_effectiveness_2024]:
\begin{equation}
\hat{S_t} \sim \text{Normal}(S_t,1/\tau_t)T(0,1).
  (\#eq:hatSEst)
\end{equation}

@@

knitr::kable(tbl,caption="Prior means and standard deviations of Gaussian distributed survival and reproduction model parameters. Standard deviations are calibrated so that the 95\\% prior predictive intervals for survival and recruitment from the Bayesian model match the 95\\% predictive intervals from Johnson et al's regression model (Supplement C).")
```

@@

As suggested by Dunham et al. [-@dunham_strategic_2023] we also calculate the expected value of perfect information (EVPI) and the expected value of sample information (EVSI) [@runge_which_2011;@yokota_value_2004]. We assign a value or utility of $\nu_{\alpha,\psi}=1$ when the true state of the population $\psi$ matches the assigned status $\alpha$, and value of $\nu_{\alpha,\psi}=0$ when the status assessment is incorrect. This is "the value ascribed by the decision-maker to the outcome, and is thus the measure of management performance" [@runge_which_2011]. The expected value of current information is the average utility of decisions informed by knowledge of the disturbance scenario $\tau$ (Fig \@ref(fig:figDisturbance)) and priors from the national model:
\begin{equation}
EV_{current} = max_{\alpha}\sum_\psi[P(\psi|\tau)\nu_{\alpha,\psi}],
  (\#eq:EVcurrent)
\end{equation}
where $P(\psi|\tau)$ is the prior probability that the true state of the population is $\psi$, and each population is assigned a status $\alpha$ informed by the prior. The expected value under certainty (i.e. the expected value of a decision when the true state of the population is known) is $EV_{certainty}=1$. The difference between $EV_{certainty}$ and $EV_{current}$ gives EVPI, a measure of the maximum possible benefit that could theoretically be gained if the system were perfectly known.

EVSI measures expected improvement from a particular sampling strategy as the difference between the expected value with and without additional sample information: 
\begin{equation}
EVSI_{\tau} = E_x[max_{\alpha}\sum_\psi[P(\psi|x,\tau)\nu_{\alpha,\psi}]] - EV_{current},
  (\#eq:EVSI)
\end{equation}
where $x$ is the sample information from a particular example trajectory. In this study, we treat simulated observations from 500 example trajectories for each disturbance and monitoring scenario $\tau$ (Figure \@ref(fig:figDisturbance)) as samples from the distribution of possible values of $x$, and assume each of these sample results is equally probable. $P(\psi|x,\tau)$ is the posterior probability that the true state of the population is $\psi$, and each population is assigned a status $\alpha$ informed by the sample information.

@@

Our analysis becomes more informative when we examine the probability of correctly assessing population status (Fig \@ref(fig:figStatusErrors)) and the expected value of information (Figs \@ref(fig:figEVSI) and E.1). When anthropogenic disturbance is high (right panels of Figs \@ref(fig:figStatusErrors), \@ref(fig:figEVSI), and E.1)), projections of population status informed only by priors from the national demographic-disturbance relationships are accurate, and adding additional local monitoring data does not improve accuracy. Interestingly, when the projection period is long and anthropogenic disturbance is high, we see a negative EVSI (lower right panel of Fig \@ref(fig:figEVSI)), and we do not see an associated decrease in the probability of correctly assessing population status (lower right panel of Fig \@ref(fig:figStatusErrors)). This means that the long-term projections informed by local monitoring data are on average less precise but not less accurate, which can happen when there is very small prior probability that population is viable (orange band in the top right panel of Fig \@ref(fig:examplePredictions) falls below 1). At lower levels of disturbance, local monitoring can reduce uncertainty, but there are diminishing returns as the number of collars deployed or the number of years of monitoring increases (Figs \@ref(fig:figStatusErrors) and \@ref(fig:figEVSI)). The shapes of the relationships between monitoring effort and the probability of correct status assessment depend on how anthropogenic disturbance is changing over time, but there are diminishing benefits of increased monitoring regardless of whether disturbance is increasing or decreasing (Supplement D). Adding unknown bias increases uncertainty, but these increases are not large for the set of parameters we investigate (Supplement F).

@@

```{r figStatusErrors, fig.cap="The probability of a correct status assessment varies with the amount of anthropogenic disturbance (columns), the number of years projected (rows), and monitoring effort (see Figure \\@ref(fig:figDisturbance) for scenarios). When disturbance is high the probability of correct status assessment is high even without local monitoring. Where disturbance is low or moderate, increasing the number of years of monitoring and the number of collars reduces uncertainty, but there are diminishing returns on additional monitoring (number of years or number of collars) as the number of years or collars increases. In these examples the composition cow multiplier $w$ is 6; varying $w$ does not alter qualitative conclusions (Supplements D and E).", cache=TRUE}
knitr::include_graphics(here::here("figs/forPaper/powercmult6ay3aSf1.png"))
```

@@

There are a variety of possible actions intended to improve survival and reproduction of populations that are not otherwise self-sustaining [@mcnay_demographic_2022; @serrouya_predicting_2020]. The simple models in this paper do not include these actions, and cannot be used to project outcomes for managed populations. Our intention is to reduce uncertainty that can be used to delay conservation action; there is nothing in our arguments, models or results that can or should be used to discourage much needed monitoring of the effects of conservation actions. Even when no additional monitoring is required for assessing self-sustainability, there are a variety of reasons to continue monitoring highly disturbed boreal caribou populations. Our point is that action is also required, and lack of monitoring should not be used as a reason to delay conservation measures for highly disturbed populations.

@@

Another novel aspect of this work is a term that allows for bias associated with misclassification of adults and missing calves in composition surveys [@ellington_divergent_2020; @decesare_estimating_2012]. Lacking information about the true magnitude of this bias, we examined sensitivity to a range of assumed values, and found that adding unknown bias did increase uncertainty, but these increases were not large (Supplement E). A key reason for this result is that we assumed the same range of values for the probability of misclassifying adults and the probability of missing calves. Presuming that surveyed groups contain more cows than bulls, and that there is an equal probability of misclassifying cows and young bulls and vice versa, then misclassification errors cause high estimated recruitment. Missing calves causes low estimated recruitment, and together these errors tend to cancel one another out. We did not investigate sensitivity to the assumption that new recruits are 50% females, but we note that this is another possible source of error that would most likely lead to low recruitment estimates.

@@

The main objective of Johnson et al's [-@johnson_science_2020;@eccc_scientific_2011] national analysis was to inform a national disturbance policy that applies to all ranges except SK1, and to inform a different policy for SK1 [@eccc_scientific_2011;@eccc_boreal_2019]. A hierarchical analysis [@leasure_hierarchical_2019] would be better suited for estimating variation in demographic model parameters among populations, and distinguishing that from variation among years. However, data owners have not authorized the use of the data for that purpose. In order to proceed with the available published information we constrained our example trajectory simulations by assuming that populations remain in their quantile, so populations that are lower than or higher than average remain lower or higher than average over time. The Bayesian model is not similarly constrained, and can accommodate evidence of a shift from unexpectedly low to unexpectedly high demographic rates over time, or vice versa. Switching from low to high each year is unlikely, but there is evidence of variation in the slope of demographic-disturbance relationships among populations [@rudolph_demographic_2017] so it is important that the Bayesian model allows for a greater range of possibilities than our more constrained simulated example trajectories. Although our simulated trajectories do not include all plausible possibilities we believe they are sufficient to demonstrate how a Bayesian model can be used to integrate local data with national priors, and to show how monitoring requirements depend on landscape condition.
